---
title: "Exploring the BRFSS data"
author: A Suarez-Pierre
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
```

### Load data

Make sure your data and R Markdown files are in the same directory. When
loaded your data file will be called `brfss2013`. Delete this note when 
before you submit your work. 

```{r load-data}
load("brfss2013.RData")
```



* * *

## Part 1: Data

This entire sample is subject to non-response bias; all the collected data
comes from people who where willing to answer the phone and finish a very 
long survey.

It is not a great representation of the general population, but the sample
size is useful to find statistical significance to even small differences.

For computing capabilities, we will refer to the data collected in the state
of California only. 

```{r subsetCA}
df = brfss2013
rm(brfss2013)                             # get rid of long name

df = df[20000:40000,]                     # trimming data, no CA rows deleted
df = tbl_df(df)         
df = filter(df, X_state=="California")    # subsetting for CA state only

dim(df)
```
Due to the fact that this has been collected retrospectively we can only
imply CORRELATIONS not CAUSALITY. 

* * *

## Part 2: Research questions

**Research quesion 1:** *What are the statistical characteristics of the 
estimaged age-gender specific maximum oxygen consumption? Does it follow a
normal distribution?*

variable: `$maxvo2_` (continuous) in ml/min/kg

**Research quesion 2:** *Does a correlation exist between the hours slept 
each night and reported days with poor physical health or poor mental health
in the last 30 days?*

explanatory variable: `sleptim1` in hours

response variables: `physhlth` and `menthlth` in days

      all of these variables are continuous

**Research quesion 3:** *Does a correlation exist between the body mass index
(BMI) of the individual answering the survey and having a high blood
cholesterol or having a heart attack?*

explanatory variable: `_bmi5` (continuous) computed by researchers in kg/m^2

response variables: `toldhi2` and `cvdinfr4`

      both response variables are discrete

* * *

## Part 3: Exploratory data analysis

---

####Research quesion 1:
*What are the statistical characteristics of the 
estimaged age-gender specific maximum oxygen consumption? Does it follow a
normal distribution?*


The last two digits of this variable are implied decimal places, before doing 
anything this corrects to the units: `ml/min/kg`

```{r maxvo2.1}
df$maxvo2_ = as.numeric(df$maxvo2_)/100
```

We will now get some descritptive statistics on the computed measurement.
```{r maxvo2.2}
summary(df$maxvo2_)
sd(df$maxvo2_, na.rm=TRUE)
```

Next some exploratory graphs
```{r maxvo2.3}
maxvo2 = ggplot(df, aes(x=maxvo2_)) + theme_bw() + 
          geom_histogram(binwidth=2, alpha=0.5, aes(y=..density..), fill="lightblue", col="grey") +
           stat_function(fun=dnorm, args=list(mean=30.31, sd=8.7), col="red") +
            geom_rug() + 
             labs(title="Distribution of VO2 max in California Population",
                  x="VO2 max (ml O2/min/kg)",
                  y="Frequency")
maxvo2
```

      The red superimposed curve is a normal distribution with the same 
      mean and standard deviation that this sample has. Nevertheless, this 
      is not the most accurate way to determine if `maxvo2` has a normal
      distribution. 

A another visual method to evaluate if this sample follow a sample
distribution is a normal probability plot using the `qqnorm` function

```{r maxvo2.4}
qqnorm(df$maxvo2_)
qqline(df$maxvo2_)
```

The sample looks like it does not follow a normal distribution, it is
seem to be ***bimodal*** as evidenced from the 'S' shape of the QQ plot.

Regardless, we can use the `Shapiro-Wilk normality test` to assess this more
thoroughly. The null hypothesis (Ho) is that the data follow a normal
distribution, and we will reject the Ho if the `p-value < 0.05`.

```{r}
shapiro.test(sample(df$maxvo2_, 5000))
```
      Unfortunately, the shapiro.test function only allows for samples
      with sizes between 3 and 5000 observations, therefore we take a 
      random sample.

**Conclusion** `p < 2.2e-16`, we can assume that this sample **does not 
follow a normal distribution.**

---

####Research quesion 2:
*Does a correlation exist between the hours slept each night and reported 
days with poor physical health or poor mental health in the last 30 days?*

Turning all the values into `numeric`
```{r sleep.1}
df$sleptim1 = as.numeric(df$sleptim1)
df$physhlth = as.numeric(df$physhlth)
df$menthlth = as.numeric(df$menthlth)
```

Descriptive statistics 
```{r sleep.2}
stats = summarize(df, "sleptim1", mean(sleptim1, na.rm=T), sd(sleptim1, na.rm=T))
colnames(stats) = c("Variable", "mean", "sd")

temp = summarize(df, "physhlth", mean(physhlth, na.rm=T), sd(physhlth, na.rm=T))
colnames(temp) = c("Variable", "mean", "sd")
stats = rbind(stats, temp)

temp = summarize(df, "menthlth", mean(physhlth, na.rm=T), sd(menthlth, na.rm=T))
colnames(temp) = c("Variable", "mean", "sd")
stats = rbind(stats, temp)

stats
```

Lets test the correlation between hours slept each night and reported days
with poor physical health in the last 30 days
```{r sleep.3}
ggplot(df, aes(x=sleptim1, y=physhlth)) +
 theme_bw() + 
  geom_jitter(alpha=0.25) + 
   geom_smooth() + 
    labs(x="Hours slept / night", y="Poor physical health (days / month)")

cor(df$sleptim1, df$physhlth, use="complete.obs")
```
There appears to be no clear correlation in the exploratory graph. Also the coefficient of correlation `r=-0.04` is not significant.

Now, lets test the correlation between hours slept each night and reported
days with poor mental health in the last 30 days
```{r sleep.4}
ggplot(df, aes(x=sleptim1, y=menthlth)) +
 theme_bw() + 
  geom_jitter(alpha=0.25) + 
   geom_smooth() + 
    labs(x="Hours slept / night", y="Poor mental health (days / month)")

cor(df$sleptim1, df$menthlth, use="complete.obs")
```
Same thing, no visual correlation or significant coefficient can be deducted
`r=-0.11`.

**Conclusion:** ***Even though there is no linear relationship between the
hours slept and both dependent variables, there is a dip in the lowess
regression line plotted which leads to infer that people on both extremes of
the spectrum might have a higher number of days with poor physical or mental
health.***


---


####Research quesion 3:
*Does a correlation exist between the body mass index (BMI) of the 
individual answering the survey and having a high blood cholesterol or 
having a heart attack?*


The last two digits of this variable are implied decimal places, before doing 
anything this corrects to the units: `kg/m^2`
```{r obesity.1}
df$X_bmi5 = as.numeric(df$X_bmi5)/100
```

Descriptive statistics
```{r obesity.2}
summary(df$X_bmi5)
sd(df$X_bmi5, na.rm=TRUE)

table(df$toldhi2) # High cholesterol?
table(df$cvdinfr4) #Ever had a heart attack?
```

Exploratory graph for high cholesterol
```{r obesity.3}
ggplot(df, aes(x=toldhi2, y=X_bmi5, fill=toldhi2)) + 
 theme_bw() + 
  geom_violin() + 
   labs(title="Violin Plot",
        x="Ever told you have high cholesterol?",
        y="BMI (kg/m^2)")
```

Statistical analysis for high cholesterol 
```{r obesity.4}
temp = filter(df, toldhi2=="Yes" | toldhi2=="No") %>%
            select(X_bmi5, toldhi2)
chisq.test(temp$X_bmi5, temp$toldhi2)
```
`p=2.9e-4` therefore we can reject the Ho that states that there is no
correlation between having a higher BMI and suffering hypercholesterolemia.

Exploratory graph for heart attack
```{r obesity.5}
ggplot(df, aes(x=cvdinfr4, y=X_bmi5, fill=cvdinfr4)) + 
 theme_bw() + 
  geom_violin() +
   labs(title="Violin Plot",
        x="Have you ever had a heart attack?",
        y="BMI (kg/m^2)")
```

Statistical analysis for heart attack
```{r obesity.6}
temp = filter(df, cvdinfr4=="Yes" | cvdinfr4=="No") %>%
            select(X_bmi5, cvdinfr4)
chisq.test(temp$X_bmi5, temp$cvdinfr4)
```
`p=6.47e-13` therefore we can reject the Ho that states that there is no
correlation between having a higher BMI and having a heart attack.


**Conclusion:** ***Having a higher BMI correlates statistically with 
having hypercholesterolemia or having a heart attack.***